---
title: "2025 Referred Isolates – Structured Survey Sampling"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

**Purpose:** Structured selection of referred bacterial isolates for AMR surveillance from the 2025 ARSRL database. For six priority organisms (ECO, KPN, ABA, PAE, EFA, SAU), the script draws the first **15 resistant/intermediate** and **5 susceptible** isolates per site, sorted by collection date. WGS results are pulled from a Google Sheet, deduplicated, and joined to flag sequencing status and concordance. Output is a date-time stamped Excel workbook with a master list, selection list, and site-level summary per organism.

---

```{r packages}
library(readxl)        # read Excel input file
library(dplyr)         # data wrangling
library(tidyr)         # pivot_wider for summary tables
library(openxlsx)      # Excel export
library(googlesheets4) # read WGS results from Google Sheets
```

---

## 1. Load Data

### 1a. Isolate database (Excel)
> Source: `Combined Data on Referred Isolates_05.14.26.xlsx`, sheet `ARSRL`.

```{r load}
df <- read_excel(
  "D:/Google Drive/My Drive/Referred_Access_db/2025/Combined Data on Referred Isolates_05.14.26.xlsx",
  sheet = "ARSRL",
  col_types = "text"   # prevent type misdetection on sparse columns
)
```

### 1d. Year filter → `df_2025`
> Filter to current year records only using `accession_no` prefix.
> Update `YEAR_PREFIX` in this chunk for succeeding years (e.g. `"26ARS"`).

```{r year_filter}
YEAR_PREFIX <- "25ARS"   # change to "26ARS", "27ARS" etc. for succeeding years

df_2025 <- df |> filter(startsWith(accession_no, YEAR_PREFIX))
cat("Records for", YEAR_PREFIX, ":", nrow(df_2025), "of", nrow(df), "total\n")
```

### 1e. Data Cleaning → `df_2025`
> All cleaning applied once to `df_2025` before any analysis.

```{r cleaning}
df_2025 <- df_2025 |>

  # Drop unused columns
  select(-any_of(c("ct/ctl", "tz/tzl", "cn/cni", "recommendation",
                   "growth", "urine_colct", "arsrl_pre", "reason",
                   "Concordance_Check"))) |>

  # Convert Excel serial numbers (stored as text) to Date
  # All date columns use origin "1899-12-30" (Excel epoch)
  mutate(
    spec_date     = as.Date(as.numeric(spec_date),     origin = "1899-12-30"),
    referral_date = as.Date(as.numeric(referral_date), origin = "1899-12-30"),
    date_birth    = as.Date(as.numeric(date_birth),    origin = "1899-12-30"),
    date_admis    = as.Date(as.numeric(date_admis),    origin = "1899-12-30")
  ) |>

  # Extract laboratory code from characters 7–9 of accession_no
  # e.g. 25ARS_BRT0034 → BRT
  mutate(laboratory = substr(accession_no, 7, 9)) |>
  relocate(laboratory, .after = accession_no) |>

  # Merge diagnosis columns: diagnosis_icd10 takes priority; falls back to diagnosis
  mutate(diagnosis = coalesce(diagnosis_icd10, diagnosis)) |>
  select(-diagnosis_icd10)
```

### 1b. WGS results (Google Sheets)
> Raw WGS run results; may contain duplicates per sample due to repeat sequencing.

```{r read_gsheet, message=FALSE, include=FALSE}
df_WGS_result <- read_sheet(
  "1NHu-RIgDeTPKgrnTTHS9qVbCe04tIeYX6k73Jm93gDw",
  sheet = "result_2025"
)
```

### 1c. Deduplicate WGS results → `df_WGS_result_dedup`
> Some samples were sequenced more than once (e.g. repeat after initial failure).
> **Dedup rules:**
> 1. PASSED > FAILED (keep the best result per sample)
> 2. If both PASSED, keep the most recent run by `batch_date`
>
> `batch_code` format: `YYYY-MM-DD_P/A/N` — suffix stripped to extract date.
> `seq_id` = combined sample + batch identifier for downstream file matching.

```{r dedup_wgs}
df_WGS_result_dedup <- df_WGS_result |>
  mutate(
    # Extract run date from batch_code (strip _P / _A / _N suffix)
    batch_date  = as.Date(sub("_[A-Z]+$", "", batch_code)),
    result_rank = case_when(
      final_result == "PASSED" ~ 1L,  # best
      final_result == "FAILED" ~ 2L,
      TRUE                     ~ 3L
    )
  ) |>
  arrange(sample_id, result_rank, desc(batch_date)) |>
  distinct(sample_id, .keep_all = TRUE) |>          # keep top row per sample
  select(-batch_date, -result_rank) |>
  mutate(
    # Sequence file ID: 25ARS_APM0010 + 2025-09-16_P → 25ARS-APM0010-20250916P
    seq_id = paste0(
      gsub("_", "-", sample_id),     # underscore → hyphen in sample ID
      "-",
      gsub("[-_]", "", batch_code)   # remove dashes/underscore from batch date
    )
  ) |>
  relocate(seq_id, .after = sample_id)

cat("WGS records after deduplication:", nrow(df_WGS_result_dedup), "\n")
```

---

## 2. Sampling Configuration

```{r config}
# Sample size per laboratory per organism
N_RI <- 15   # Resistant or Intermediate
N_S  <- 5    # Susceptible

# Resistance markers — both disk (ND) and MIC (NM) interpretation columns included.
# An isolate is flagged if ANY column in the set = "R" or "I".

# Carbapenems: IPM, MEM, ETP, DOR  →  ECO, KPN, ABA, PAE
CARBAPENEM_RIS <- c(
  "ipm_nd10_ris", "ipm_nm_ris",
  "mem_nd10_ris", "mem_nm_ris",
  "etp_nd10_ris", "etp_nm_ris",
  "dor_nd10_ris", "dor_nm_ris"
)

# Vancomycin  →  EFA (VRE screen)
VAN_RIS <- c("van_nd30_ris", "van_nm_ris")

# Oxacillin / Cefoxitin  →  SAU (MRSA screen)
MRSA_RIS <- c("oxa_nd1_ris", "oxa_nm_ris", "fox_nd30_ris", "fox_nm_ris")

# WGS concordance: expected genus per organismcode.
# Genus-level match is accepted (e.g. "Klebsiella" matches kpn).
# Exception: Enterococcus faecium is DISCORDANT for efa.
ORG_GENUS_MAP <- c(
  eco = "Escherichia",
  kpn = "Klebsiella",
  aba = "Acinetobacter",
  pae = "Pseudomonas",
  efa = "Enterococcus",
  sau = "Staphylococcus"
)

# ── Specimen type filters ─────────────────────────────────────────────────────
# Shared base sets — defined once, reused across organisms.

SPEC_URINE_ALL <- c("uc", "ur", "cv", "ub", "uk", "ua", "fv", "uz", "uo", "uu")

SPEC_TISSUE_ALL <- c(
  "ea", "ey", "mo", "sk", "jt", "og", "um", "bx", "ti", "bn",
  "he", "hv", "li", "mu", "ha", "ar", "ax", "co", "em", "eo",
  "fi", "pl", "bt", "bo", "bm", "ki", "du", "me", "bu", "cy",
  "ln", "gl", "nl", "ft", "au", "us", "cr", "pn", "tn", "ap",
  "cb", "ec", "ed", "eg", "eh", "fa", "gb", "hi", "kn", "lx",
  "lp", "lo", "ov", "pi", "po", "rc", "sl", "te", "to", "bv",
  "hd", "sv", "lg", "nk", "nv", "lv", "gr", "ig"
)

SPEC_RESP_KPN <- c("no", "th", "sp", "br", "lu", "tr", "si",
                   "ba", "rl", "ru", "ta", "rp", "np", "sa", "in")

# Per-organism allowlists (update here only if spec eligibility changes)
SPEC_TYPE_ECO <- c("bl", "sf", SPEC_URINE_ALL)                                  # Blood + CSF + all Urine
SPEC_TYPE_KPN <- c("bl", "sf", SPEC_URINE_ALL, SPEC_TISSUE_ALL, SPEC_RESP_KPN)  # Blood + CSF + all Urine + all Tissue + all Respiratory
SPEC_TYPE_ABA <- c("bl", "sf", "ta", SPEC_URINE_ALL, SPEC_TISSUE_ALL)           # Blood + CSF + ta only + all Urine + all Tissue
SPEC_TYPE_PAE <- SPEC_TYPE_KPN                                                   # identical to KPN
SPEC_TYPE_EFA <- c("bl", "sf", SPEC_URINE_ALL)                                  # Blood + CSF + all Urine
SPEC_TYPE_SAU <- SPEC_TYPE_ABA                                                   # identical to ABA
```

---

## 3. Helper Functions

```{r helpers}
# ── attach_wgs() ──────────────────────────────────────────────────────────────
# Joins WGS data (final_result, seq_id, wgs_id) to isolate records via AccessionNo.
# Steps:
#   1. Left-join df_WGS_result_dedup on AccessionNo = sample_id
#   2. Concordance check — genus of wgs_id vs expected genus for that organism
#      → DISCORDANT overrides final_result to "FAILED"
#   3. Derive WGS_Status: PASSED / FAILED / No (not sequenced)
#   4. Derive WGS_Status label (PASSED / FAILED / No) for reporting only — no effect on sort order
attach_wgs <- function(data) {
  data |>
    left_join(
      df_WGS_result_dedup |> select(sample_id, final_result, seq_id, wgs_id),
      by = c("accession_no" = "sample_id")
    ) |>
    mutate(
      wgs_genus      = sub(" .*$", "", wgs_id),          # first word = genus
      expected_genus = ORG_GENUS_MAP[organismcode],
      concordance    = case_when(
        is.na(wgs_id)                                                         ~ NA_character_,
        wgs_genus == expected_genus &
          !(organismcode == "efa" & wgs_id == "Enterococcus faecium")         ~ "CONCORDANT",
        TRUE                                                                  ~ "DISCORDANT"
      ),
      final_result   = if_else(concordance == "DISCORDANT", "FAILED", final_result),
      WGS_Status     = case_when(
        final_result == "PASSED" ~ "PASSED",
        final_result == "FAILED" ~ "FAILED",
        TRUE                     ~ "No"
      ),
    ) |>
    select(-wgs_genus, -expected_genus)
}

# ── sample_org() ──────────────────────────────────────────────────────────────
# For a single organism's data:
#   - Attaches WGS info (attach_wgs)
#   - Selects first N_RI R/I isolates per laboratory, sorted by spec_date (earliest first)
#   - Selects first N_S   S   isolates per laboratory, same sort
# Returns combined list with a Category column.
sample_org <- function(data, ris_cols, n_ri, n_s) {
  data <- attach_wgs(data)

  ri <- data |>
    filter(if_any(any_of(ris_cols), ~ . %in% c("R", "I"))) |>
    arrange(laboratory, spec_date) |>
    group_by(laboratory) |>
    slice_head(n = n_ri) |>
    ungroup() |>
    mutate(Category = "Resistant/Intermediate")

  sus <- data |>
    filter(if_all(any_of(ris_cols), ~ is.na(.) | . == "S")) |>
    arrange(laboratory, spec_date) |>
    group_by(laboratory) |>
    slice_head(n = n_s) |>
    ungroup() |>
    mutate(Category = "Susceptible")

  bind_rows(ri, sus)
}

# ── drop_empty_cols() ─────────────────────────────────────────────────────────
# Removes columns where every value is NA (antibiotics not tested for that organism).
drop_empty_cols <- function(data) {
  data[, colSums(!is.na(data)) > 0, drop = FALSE]
}

# ── summarise_org() ───────────────────────────────────────────────────────────
# Site-level distribution table:
#   Columns: laboratory | R/I count | S count | Total | RI_WGS_Passed | S_WGS_Passed | Total_WGS_Passed
#   Last row: column totals (TOTAL).
# WGS_Passed counts only PASSED — discordant isolates are already set to FAILED
# upstream so they are excluded automatically.
summarise_org <- function(sample_data) {
  counts <- sample_data |>
    group_by(laboratory, Category) |>
    summarise(Count = n(), .groups = "drop") |>
    complete(laboratory, Category = c("Resistant/Intermediate", "Susceptible"),
             fill = list(Count = 0)) |>
    pivot_wider(names_from = Category, values_from = Count, values_fill = 0)

  ri_wgs <- sample_data |>
    filter(Category == "Resistant/Intermediate") |>
    group_by(laboratory) |>
    summarise(RI_WGS_Passed = sum(WGS_Status == "PASSED"), .groups = "drop")

  s_wgs <- sample_data |>
    filter(Category == "Susceptible") |>
    group_by(laboratory) |>
    summarise(S_WGS_Passed = sum(WGS_Status == "PASSED"), .groups = "drop")

  tbl <- counts |>
    left_join(ri_wgs, by = "laboratory") |>
    left_join(s_wgs,  by = "laboratory") |>
    replace_na(list(RI_WGS_Passed = 0L, S_WGS_Passed = 0L)) |>
    mutate(
      Total            = rowSums(across(c(`Resistant/Intermediate`, Susceptible))),
      Total_WGS_Passed = RI_WGS_Passed + S_WGS_Passed
    ) |>
    select(laboratory, `Resistant/Intermediate`, Susceptible, Total,
           RI_WGS_Passed, S_WGS_Passed, Total_WGS_Passed) |>
    arrange(laboratory)

  total_row <- tbl |>
    summarise(across(where(is.numeric), sum)) |>
    mutate(laboratory = "TOTAL")

  bind_rows(tbl, total_row)
}
```

---

## 4. Save WGS Data

> Raw and deduplicated WGS results saved as separate Excel files for audit/reference.

```{r save_wgs}
stamp <- format(Sys.Date(), "%Y-%m-%d")

# Add seq_id to raw (all rows, duplicates included); same formula as dedup
wgs_raw_export <- df_WGS_result |>
  mutate(seq_id = paste0(gsub("_", "-", sample_id), "-", gsub("[-_]", "", batch_code))) |>
  relocate(seq_id, .after = sample_id)

# One workbook, two sheets: raw (with seq_id) and deduplicated (seq_id already present)
wb_wgs <- createWorkbook()
addWorksheet(wb_wgs, "WGS_Raw");   writeData(wb_wgs, "WGS_Raw",   wgs_raw_export)
addWorksheet(wb_wgs, "WGS_Dedup"); writeData(wb_wgs, "WGS_Dedup", df_WGS_result_dedup)

saveWorkbook(wb_wgs,
  paste0("D:/R/structured/selection/output/WGS_results_", stamp, ".xlsx"),
  overwrite = TRUE)
cat("Saved: WGS_results_", stamp, ".xlsx\n", sep = "")
```

---

## 5. Data Preparation & Sampling

```{r prep}
# df_2025 is cleaned and ready — see cleaning chunk in Section 1e
```

```{r sampling}
# Each organism filtered from df_2025 (organismcode + spec_type), then passed through sample_org()
# Resistance screen:  ECO/KPN/ABA/PAE → carbapenems | EFA → VAN | SAU → OXA/FOX
# Spec-type allowlists defined in config chunk (SPEC_TYPE_*) — update there, not here.
eco_sample <- df_2025 |> filter(organismcode == "eco", spec_type %in% SPEC_TYPE_ECO) |> sample_org(CARBAPENEM_RIS, N_RI, N_S)
kpn_sample <- df_2025 |> filter(organismcode == "kpn", spec_type %in% SPEC_TYPE_KPN) |> sample_org(CARBAPENEM_RIS, N_RI, N_S)
aba_sample <- df_2025 |> filter(organismcode == "aba", spec_type %in% SPEC_TYPE_ABA) |> sample_org(CARBAPENEM_RIS, N_RI, N_S)
pae_sample <- df_2025 |> filter(organismcode == "pae", spec_type %in% SPEC_TYPE_PAE) |> sample_org(CARBAPENEM_RIS, N_RI, N_S)
efa_sample <- df_2025 |> filter(organismcode == "efa", spec_type %in% SPEC_TYPE_EFA) |> sample_org(VAN_RIS,        N_RI, N_S)
sau_sample <- df_2025 |> filter(organismcode == "sau", spec_type %in% SPEC_TYPE_SAU) |> sample_org(MRSA_RIS,       N_RI, N_S)

samples <- list(
  ECO = eco_sample, KPN = kpn_sample, ABA = aba_sample,
  PAE = pae_sample, EFA = efa_sample, SAU = sau_sample
)

# Console summary: selection counts + WGS breakdown per organism
for (org in names(samples)) {
  d <- samples[[org]]
  cat(org,
      "— RI:",         sum(d$Category   == "Resistant/Intermediate"),
      "| S:",          sum(d$Category   == "Susceptible"),
      "| WGS PASSED:", sum(d$WGS_Status == "PASSED"),
      "| WGS FAILED:", sum(d$WGS_Status == "FAILED"),
      "| No WGS:",     sum(d$WGS_Status == "No"), "\n")
}
```

---

## 6. Export to Excel

> One workbook per run, filename date-time stamped.
> **Sheet order per organism:** `{ORG}_Master` → `{ORG}` (selection) → `{ORG}_Summary`
> Name columns (first_name, mid_name, last_name) excluded from all sheets.
> All-NA columns dropped per organism (antibiotics not applicable to that species).

```{r export}
wb <- createWorkbook()

for (org in names(samples)) {

  # Master: full organism records from df_2025 with WGS info attached
  master <- df_2025 |>
    filter(organismcode == tolower(org)) |>
    attach_wgs() |>
    select(-any_of(c("first_name", "mid_name", "last_name"))) |>
    drop_empty_cols()

  # Selection: 15 RI + 5 S per laboratory, sorted by spec_date
  data <- samples[[org]] |>
    select(-any_of(c("first_name", "mid_name", "last_name"))) |>
    drop_empty_cols()

  # Summary: site × category counts with WGS_Passed breakdown
  summary <- summarise_org(samples[[org]])

  addWorksheet(wb, paste0(org, "_Master"));  writeData(wb, paste0(org, "_Master"),  master)
  addWorksheet(wb, org);                     writeData(wb, org,                     data)
  addWorksheet(wb, paste0(org, "_Summary")); writeData(wb, paste0(org, "_Summary"), summary)
}

out_file <- paste0(
  "D:/R/structured/selection/output/referred_sample_",
  format(Sys.time(), "%Y-%m-%d_%H%M%S"),
  ".xlsx"
)
saveWorkbook(wb, out_file, overwrite = TRUE)
cat("Saved:", out_file, "\n")
```
